\large \textbf{Leveraging Historical Ties between Cognitive Science and Computer Science to Guide Innovative Use of Computers in Education}

### 2016-02-01 paper.tex
tweaks and comments after chat with Jake

% sorting=none will order citations in order of appearance
% use sorting=nyt to sort by name, year, title
\usepackage[backend=biber,autolang=hyphen, bibencoding=inputenc, isbn=false, uniquename=false, sorting=none, sortcites=true, style=numeric, autocite=superscript, url=false]{biblatex} % biblatex setup
\DeclareLanguageMapping{american}{american-apa}
\appto{\bibfont}{\footnotesize} % for 10pt bibliography
%\usepackage{xunicode}
\usepackage[american]{babel}
\usepackage[babel]{csquotes}
\usepackage[utf8]{inputenc}
\usepackage{float}
\usepackage{graphicx}
\usepackage[margin=1in]{geometry}
\usepackage[dvipsnames,table]{xcolor}
\usepackage{tikz}
\usetikzlibrary{positioning,arrows,shapes,fit}
\usepackage{standalone}
\usepackage{mdframed}
\usepackage{capt-of}
\usepackage{changepage}
\usepackage{titlesec}
\usepackage[parfill]{parskip}

@@

Juxtaposed with the observation that technology is changing rapidly,
and the half-life of technical skills is decreasing (cite half-life
article) it would seem teaching programming is a questionable use of
class time: the language and paradigm students are taught will likely
not be the same they are asked to use in their first job
post-graduation. However, if we acknowledge that learning to program
is more than simply learning to write code, and in particular if we
draw on work from the early era of computing which closely linked
programming to problem solving and cognitive skills, we can argue that
regardless of language or paradigm, learning to program may actually
be a way to learn to problem solve, reason about and organize
knowledge, and conceptualize processes.

While there are some recent studies which investigate programming
education as a means of teaching problem-solving, there is also a
large body of research that primarily focuses on teaching students the
details of how to code. We argue that in most cases these details are
less important than centering programming education around
problem-solving skills and if educators hope to achieve this goal then
all aspects of introductory programming courses, from the course
objectives, to the curriculum used, to the assessments administered,
must be chosen intentionally with this goal in mind.

@@

\section{Methods}
Assuming that there would be more comprehensive research of
introductory programming in the computer science community we began
with targeted searchs of the ACM digital library for `introductory
programming'. Once we found key papers that seemed to indicate a call
to action
\autocite[e.g.][]{mccracken_multinational_2001,kolling_problem_1999},
or a survey of practices
\autocite[e.g.][]{pears_survey_2007-1,robins_learning_2003} we used
the `cited by' list to find additional relevant papers.

@@

\section{Current state of research}
The bulk of literature about teaching introductory programming (CS1)
revolves around what paradigms are “best” (e.g. procedural vs. object
oriented), which language is “best”, often times with regard to a
particular paradigm (e.g. OOP with Python vs. Java), and best teaching
practices (e.g. use of “pair programming” environments). However, the
\emph{reasons} for picking one paradigm over another tend to be about
industry demand, rather than intentionally choosing a paradigm to
support the learning objectives of the course.

@@

The \textcite{mccracken_multinational_2001} study, which concluded that
regardless of nation or institution, most students do not know how to
program by the end of their introductory programming courses, sparked
much conversation and push to reform introductory programming in the
next decade. Ten years after the
\citeauthor{mccracken_multinational_2001} study,
\textcite{lister_ten_2011} reflected on the response and lemented that
while much of the criticism that the problems used to test student
knowledge were too difficult, no one had repeated the study with
problems they deemed to be more reasonable. Thus while the
\citeauthor{mccracken_multinational_2001} study validated the
experiences of many introductory programming instructors it did not
result in critical change to address the problems. Rather, what we
have seen since the 1990s is been an increasing interest in where
object oriented programming should be placed within the curriculum
with debates between objects-first vs procedural-fist but little
question of why object oriented design should be included at all other
than it continues to be a popular paradigm in industry.

@@

\citeauthor{pears_survey_2007-1} conclude their
\citeyear{pears_survey_2007-1} literature survey with the observation
that despite the volume of work exploring the implementation of
introductory programming courses “there is little systematic evidence
to support any particular approach.”\autocite{pears_survey_2007-1} This
could be due to two reasons:

@@

@@

In one of these studies, the authors draw from three concepts of
cognitive science: cognitive load theory, cognitive apprenticeship,
and the theory of worked examples, to design an introductory
programming course\autocite{caspersen_instructional_2007-1}. It is
interesting to note that they select these theories based on a human
cognition model that resulted from computing: separate working and
long term memory where learning is the process for transferming
information from working to long-term memory. The course the authors
describe teachs OOP but they never explain why the chose OOP as the
paradigm, it is assumed that this is just a given feature of an
introductory programming course.

>>>>

% sorting=none will order citations in order of appearance
% use sorting=nyt to sort by name, year, title
\usepackage[backend=biber,autolang=hyphen, bibencoding=inputenc, isbn=false, uniquename=false, sorting=none, sortcites=true, style=numeric, autocite=superscript, url=false]{biblatex} % biblatex setup
\DeclareLanguageMapping{american}{american-apa}
\appto{\bibfont}{\footnotesize} % for 10pt bibliography
%\usepackage{xunicode}
\usepackage[american]{babel}
\usepackage[babel]{csquotes}
\usepackage[utf8]{inputenc}
\usepackage{float}
\usepackage{graphicx}
\usepackage[margin=1in]{geometry}
\usepackage[dvipsnames,table]{xcolor}
\usepackage{tikz}
\usetikzlibrary{positioning,arrows,shapes,fit}
\usepackage{standalone}
%\usepackage{mdframed}
\usepackage{changepage}
\usepackage{titlesec}
\usepackage[parfill]{parskip}

@@

Juxtaposed with the observation that technology is changing rapidly,
and the half-life of technical skills is
decreasing\autocite{nae_educating_2004} it would seem teaching
programming is a questionable use of class time: the language and
paradigm students are taught will likely not be the same they are
asked to use in their first job post-graduation. However, if we
acknowledge that learning to program is more than simply learning to
write code, and in particular if we draw on work from the early era of
computing which closely linked programming to problem solving and
cognitive skills, we can argue that regardless of language or
paradigm, learning to program may actually be a way to learn to
problem solve, reason about and organize knowledge, and conceptualize
processes.

While there are some recent studies which investigate programming
education as a means of teaching problem-solving, there is also a
large body of research that primarily focuses on teaching students the
details of how to code. We argue that in most cases these details are
less important than centering programming education around
problem-solving skills and if educators hope to achieve this goal then
all aspects of introductory programming courses, from the course
objectives, to the curriculum used, to the assessments administered,
must be chosen intentionally with this goal in mind. In this paper we
will present a brief summary of current published work on introductory programming education, then note a potential lack that, 
then offer a potential framework to

@@

\section{Methods}
[identify learning sciences sources]

Assuming that there would be more comprehensive research of
introductory programming in the computer science community we began
with targeted searchs of the ACM digital library for `introductory
programming'. Once we found key papers that seemed to indicate a call
to action
\autocite[e.g.][]{mccracken_multinational_2001,kolling_problem_1999},
or a survey of practices
\autocite[e.g.][]{pears_survey_2007-1,robins_learning_2003} we used
the `cited by' list to find additional relevant papers.

@@

\section{Current state of research}
%%TODO: add citations
The bulk of literature about teaching introductory programming (CS1)
revolves around what paradigms are ``best'' (e.g. procedural vs. object
oriented), which language is ``best'', often times with regard to a
particular paradigm (e.g. OOP with Python vs. Java), and best teaching
practices (e.g. use of “pair programming” environments). However, the
\emph{reasons} for picking one paradigm over another tend to be about
industry demand, rather than intentionally choosing a paradigm to
support the learning objectives of the course.
%% TODO: reminder to connect to learning science, e.g. this behavior
%% is not uncommong, for instance flipped classrooms is a ``hot
%% topic'' now and often times we see people implementing a ``flipped
%% classroom'' without necessarily considering if that pedagogical
%% approach makes the most sense for a given course.

@@

The \textcite{mccracken_multinational_2001} study, which concluded that
regardless of nation or institution, most students do not know how to
program by the end of their introductory programming courses, sparked
much conversation and push to reform introductory programming in the
next decade. Ten years after the
\citeauthor{mccracken_multinational_2001} study,
\textcite{lister_ten_2011} reflected on the response and lamented that
while much of the criticism that the problems used to test student
knowledge were too difficult, no one had repeated the study with
problems they deemed to be more reasonable. 
%% TODO: connect mccracken to engineering ed about missconceptions
%% (CHEER chapter and Streveler paper)

Thus while the \citeauthor{mccracken_multinational_2001} study
validated the experiences of many introductory programming instructors
it did not result in critical change to address the problems. Rather,
what we have seen since the 1990s is been an increasing interest in
where object oriented programming should be placed within the
curriculum with debates between objects-first vs procedural-fist but
little question of why object oriented design should be included at
all other than it continues to be a popular paradigm in industry.

@@

\citeauthor{pears_survey_2007-1} conclude their
\citeyear{pears_survey_2007-1} literature survey with the observation
that despite the volume of work exploring the implementation of
introductory programming courses “there is little systematic evidence
to support any particular approach.”\autocite{pears_survey_2007-1}
We think this could be due to two reasons:

@@

%%TODO: link to knowledge/skills interaction literature
%%TODO: bring in knowledge organization, focus on concepts vs. procedures

@@

In one of these studies, the authors draw from three concepts of
cognitive science: cognitive load theory, cognitive apprenticeship,
and the theory of worked examples, to design an introductory
programming course\autocite{caspersen_instructional_2007-1}. It is
interesting to note that they select these theories based on a human
cognition model that resulted from computing: separate working and
long term memory where learning is the process for transforming
information from working to long-term memory. The course the authors
describe teachs OOP but they never explain why the chose OOP as the
paradigm, it is assumed that this is just a given feature of an
introductory programming course.

%%TODO: consider focusing on this and dropping threshold concepts and others. Potentally bring in pattern-based pedagogy
